|

### 2019-03-22 Paper.Rmd
DONE

abstract: "This paper evaluates San Francisco's Inclusionary Housing Program by estimating the impact of affordable developments on assessed property values in the surrounding area. We use a panel dataset constructed from parcel-level municipal tax assessments and data on new residential developments. Exploiting the city's mandate that all new developments either build affordable units or pursue an alternative option such as paying an in-lieu fee, we use a difference-in-differences design to compare changes in surrounding property values between these two groups in the years before and after project completion. We find that on average, property values increased more in areas surrounding affordable developments compared to areas surrounding non-affordable developments."

@@

In this paper, we estimate the effect of San Francisco's Inclusionary Housing Program on residential property values in the areas surrounding affordable developments. Using a difference-in-differences model on a panel dataset of all eligible properties in the city, we find significant evidence to support the conclusion that the program has a positive effect on nearby property values. These results are consistent with economic theory and the existing academic literature.

@@

If the developer chooses the first option, the actual number of affordable units and levels of "affordability" required for the project are determined by a complex set of guidelines determined by the city's Planning Department. If not, the development's "in-lieu" fee is calculated on a individual basis based on property and area-specific characteristics as part of the project's approval process. For the purposes of this paper, we simplify the situation and consider developments that pursue the first option to be "affordable developments" built at the location of interest, and all others to fall into an "otherwise" category. While we would like to further study outcomes related to fee payment and off-site below-market rate construction, we could not find publicly available data regarding the size of the fee or the characteristics of off-site projects.

@@

The City of San Francisco itself conducted an analysis of the viability of its Inclusionary Housing Program in 2016. This paper, drafted by an external consulting firm, included recommendations for affordable housing policy that would have a minimal impact on the feasibility of new developments. It predicted that a 1% increase in the city's inclusionary requirement (the percentage of units in developments mandated to be designated as affordable) would decrease the supply of housing in the city by 1.8%, and increase housing prices and rents by 2%. The study also estimated the direct impact of the program on low-income households to be \$660 million, although the secondary impact of cheap housing being freed up by tenants moving into new affordable units was not taken into account. While these results are consistent with other literature on this topic, the study was primarily a simulation exercise and did not incorporate real-world evidence of the policy's effect.

@@

The identifying assumption of this model is that the pre-treatment trends in the surrounding property values for affordable and non-affordable developments are parallel, conditional on the observables that determine whether or not a developer adheres to the below-market-rate mandate. We assume that in the absence of the Inclusionary Housing Program, the trends in property values in the areas surrounding the affordable and non-affordable developments with similar characteristics would have continued in a parallel manner in the post-construction period.

@@

In our analysis below, we provide evidence to suggest that the parallel trends assumption can be justified in our data. We run several models using various combinations of controls, including a set of neighborhood indicators, such that our estimates measure the effect of the program conditional on properties having similar underlying characteristics. With this specification, we can therefore interpret the difference-in-differences estimate as the differential change in the median of *immediately* neighboring property values for treated vs. non-treated developments, given that the properties are in the same *general* neighborhood of the city. We evaluate this pre-trends assumption in the data both visually and empirically using a lagged variable model.

@@

We then calculated the distance of every development to every unique property coordinate within the tax assessment data using the great-circle distance as calculated by the Haversine formula in R's `geosphere` package. To speed up our calculation, we only examined the properties in the same neighborhood as the development, which could lead to potential inaccuracies for properties on the edge of neighborhood boundaries. However, because the distance rings are extremely small compared to the size of the city's neighborhood geographis, we imagine that these instances will be rare and unlikely to significantly affect our results.

@@

$$\text{Y}_{it}=\alpha+\beta S_{i}+\gamma T_{t}+\delta(S_{i}\times T_{t})+\rho\mathbf{X}_{it}+\epsilon_{it}$$

$Y_{it}$ is the median property value for the surrounding area $i$ in year $t$, where surrounding area is defined by a given Haversine distance band of 100, 200, or 300 meters. $S_i$ is a dummy for a simplified version of the developer's Section 415 declaration, which takes the value 1 if the developer builds below market-rate housing at the location specified in the data, and 0 otherwise (which encompasses all other available options under Section 415). $T_t$ is a dummy for whether the specified year was before (=0) or after (=1) construction of the property was completed. $\mathbf{X}_{it}$ is a vector of covariates, which includes neighborhood and property-specific controls such as .

@@

This regression is run on the merged affordable housing and tax assessment datasets for periods $t-1$ to $t+1$, where each row represents a property eligible. The difference-in-differences estimate from this regression can be interpreted as follows:

$$\hat{\delta}_{DD}=\bar{Y}_{\text{post}}^{\text{affordable}}-\bar{Y}_{\text{pre}}^{\text{affordable}}-(\bar{Y}_{\text{post}}^{\text{pay fee}}-\bar{Y}_{\text{pre}}^{\text{pay fee}})$$

@@

The coefficients are also easily interpretable: $\alpha$ is the average pre-construction median surrounding property value for non-affordable properties,  $\beta$ is the affordable housing-specific effect on surrounding property values, $\gamma$ is the time trend common to both affordable and non-affordable properties, and $\delta$ is Diff-in-diff estimator: true effect of treatment.

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}

@@

In these plots, we generally see visual evidence of the parallel trends assumption in the data as evidenced by the parallel slopes in the treatment . While there may be enough evidence to support the parallel trends assumption . We also

m1 <- lm(log.dist_100_median ~ treated + time + did + Assessor.Neighborhood.District, data = time.subset)
m2 <- lm(log.dist_200_median ~ treated + time + did + Assessor.Neighborhood.District, data = time.subset)
m3 <- lm(log.dist_300_median ~ treated + time + did + Assessor.Neighborhood.District + Number.of.Bedrooms, data = time.subset)

m4 <- lm(log.dist_100_mean ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m5 <- lm(log.dist_200_mean ~ treated + time + did, data = time.subset)
m6 <- lm(log.dist_300_mean ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)

stargazer(m1, m2, m3, type='latex', 
          dep.var.labels=c("Log(100m Median)","Log(200m Median)", "Log(300m Median)"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = c("Assessor.Neighborhood.District", "Number.of.Bedrooms"), omit.labels = c("Neighborhood dummies?"))

stargazer(m4, m5, m6, type='latex', 
          dep.var.labels=c("Log(100m Mean)","Log(200m Mean)", "Log(300m Mean)"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = "Assessor.Neighborhood.District", omit.labels = c("Neighborhood dummies?"))

# Conclusion

By leveraging the unique legal framework of San Francisco's affordable housing mandate, as well as the public availability of detailed micro-level spatial data, we evaluate the impact of the city's Inclusionary Housing Program on residential property values in the surrounding area. We find modest evidence to suggest that this policy has a positive effect on nearby home values, contradicting its supposed aim as a tool for reducing the cost of housing.

@@

While our model's estimates are highly significant within the scope of our data, we are less certain of their generalizability to other contexts. Other cities have different, and could. However, we have also reason to believe that the market dynamics within narrow geographic are . When considering the

A more comprehensive evaluation of the effectiveness of this program might also address questions related to its effect on outcomes for individuals who are selected to participate in the program. Although individual-level data is not publicly available, the random assignment generated by the program's lottery structure could allow for a rigorous quasi-experimental approach.

>>>>

abstract: "This paper evaluates San Francisco's Inclusionary Housing Program by estimating the impact of affordable developments on assessed property values in the surrounding area. We use a panel dataset constructed from parcel-level municipal tax assessments and data on new residential developments. Exploiting the city's mandate that all new developments either build affordable units or pursue an alternative option such as paying an in-lieu fee, we use a difference-in-differences design to compare changes in surrounding property values between these two groups in the years before and after project completion. We find that on average, property values increased more in areas surrounding affordable developments compared to areas surrounding non-affordable developments. In addition, we find that the magnitude of this treatment effect is negatively associated with distance."

@@

In this paper, we estimate the effect of San Francisco's Inclusionary Housing Program on residential property values in the areas surrounding affordable developments. Using a difference-in-differences model on a panel dataset of all eligible properties in the city, we find significant evidence to support the conclusion that the program has a positive effect on nearby property values, and that the magnitude of this effect decreases with distance. These results are consistent with economic theory and the existing academic literature.

@@

If the developer chooses the first option, the actual number of affordable units and levels of "affordability" required for the project are determined by a complex set of guidelines set by the city's Planning Department. If not, the development's "in-lieu" fee is calculated on a individual basis based on property and area-specific characteristics as part of the project's approval process. For the purposes of this paper, we simplify the situation and consider developments that pursue the first option to be "affordable developments" built at the location of interest, and all others to fall into an "otherwise" category. While we would like to further study outcomes related to fee payment and off-site below-market rate construction, we could not find publicly available data regarding the size of the fee or the characteristics of off-site projects.

@@

The City of San Francisco itself conducted an analysis of the viability of its Inclusionary Housing Program in 2016 [@sanfranciscoofficeofthecontrollerInclusionaryHousingWorking2016]. Their report, drafted by an external consulting firm, included recommendations for affordable housing policy that would have a minimal impact on the feasibility of new developments. It predicted that a 1% increase in the city's inclusionary requirement (the percentage of units in developments mandated to be designated as affordable) would decrease the supply of housing in the city by 1.8%, and increase housing prices and rents by 2%. The study also estimated the direct impact of the program on low-income households to be \$660 million, although the secondary impact of cheap housing being freed up by tenants moving into new affordable units was not taken into account. While these results are consistent with other literature on this topic, the study was primarily a simulation exercise and did not incorporate real-world evidence of the policy's effect.

@@

The identifying assumption of this model is that the pre-construction trends in the surrounding property values for affordable and non-affordable developments are parallel, conditional on the observables that determine whether or not a developer adheres to the below-market-rate mandate. We assume that in the absence of the Inclusionary Housing Program, the trends in property values in the areas surrounding the affordable and non-affordable developments with similar characteristics would have continued in a parallel manner in the post-construction period.

@@

In our analysis below, we provide evidence to suggest that the parallel trends assumption can be justified in our data. We run several models using various combinations of controls, including a set of neighborhood indicators, such that our estimates measure the effect of the program conditional on properties having similar underlying characteristics. With this specification, we can therefore interpret the difference-in-differences estimate as the differential change in the median of *immediately* neighboring property values for treated vs. non-treated developments, given that the properties are in the same *general* neighborhood of the city. We evaluate this pre-trends assumption in the data both visually and empirically.

@@

We then calculated the distance of every development to every unique property coordinate within the tax assessment data using the great-circle distance as calculated by the Haversine formula in R's `geosphere` package. To speed up our calculation, we only examined the properties in the same neighborhood as the development, which could lead to potential inaccuracies for properties on the edge of neighborhood boundaries. However, because the distance rings are extremely small compared to the size of the city's neighborhood geographies, we imagine that these instances will be rare and unlikely to significantly affect our results.

@@

$$\text{log(Y}_{it})=\alpha+\beta S_{i}+\gamma T_{t}+\delta(S_{i}\times T_{t})+\rho\mathbf{X}_{it}+\epsilon_{it}$$

$Y_{it}$ is the median property value for the surrounding area $i$ in year $t$, where surrounding area is defined by a given Haversine distance band of 100, 200, or 300 meters. We chose to log-transform the outcome variable $Y_{it}$ due to the presence of outliers, as mentioned earlier, which led to errors that were non-normally distributed. $S_i$ is a dummy for a simplified version of the developer's Section 415 declaration, which takes the value 1 if the developer builds below market-rate housing at the location specified in the data, and 0 otherwise (which encompasses all other available options under Section 415). $T_t$ is a dummy for whether the specified year $t$ was before (=0) or after (=1) construction of the property was completed. $\mathbf{X}_{it}$ is a vector of covariates, which includes neighborhood and property-specific controls such as Number of Bedrooms and Square Footage.

@@

The coefficients are also easily interpretable: $\alpha$ is the average pre-construction log(median) surrounding property value for non-affordable properties, $\beta$ is the affordable housing-specific effect on surrounding property values, $\gamma$ is the time trend common to both affordable and non-affordable properties, and $\delta$ is the diff-in-diff estimator for the model. As mentioned earlier, this represents the differential change in the log(median) of immediately surrounding property values for affordable vs. non-affordable developments, given that the properties are in the same neighborhood of the city and share the same underlying characteristics.

@@

We choose to include covariates in our model to further validate the parallel trend assumption underlying the diff-in-diff model. We believe we can more credibly make the assumption that the pre-construction trends in the outcome are parallel once we control for broader neighborhood and property-level factors. Therefore, all of our estimates can be interpreted as the traditional diff-in-diff coefficients for two similar developments in the same neighborhood, where one of the developments designated affordable units and the other did not. This conditional parallel trends assumption is as follows:

$$E[Y_{i1}(0) - Y_{i0}(0)|S_i = 1, \mathbf{X}_i = x] = E[Y_{i1}(0) - Y_{i0}(0)|S_i = 0, \mathbf{X}_i = x]$$

Under this specification, the average treatment effect on the treated can therefore be identified as follows:

\begin{equation*}
\begin{split}
\tau_{ATT}=\sum_x[\{\mathbb{E}[Y_{i1}|S_i=1,\mathbf{X}_i=x]-\mathbb{E}[Y_{i1}|S_i=0,\mathbf{X}_i=x]\}\\
-\{\mathbb{E}[Y_{i1}|S_i=1,\mathbf{X}_i=x]-\mathbb{E}[Y_{i1}|S_i=0,\mathbf{X}_i=x]\}]Pr(\mathbf{X}_i=x|G_i=1)\\
\end{split}
\end{equation*}

(Full derivation provided [here](http://web.mit.edu/teppei/www/teaching/Keio2016/06dd.pdf))

This regression is run on the merged affordable housing and tax assessment datasets, where each row represents a property with a mandated Section 415 requirement at year $t$. Therefore, each property corresponds to three rows, one for each of the years immediately before, during, and immediately after the construction of a given unit. We provide the regression tables for the results of the model according to several different specifications.

In Table 2, each of three columns corresponds to the dependent variable Log(Median Property Value) for properties within the 100m, 200m, and 300m distance bands, controlling for neighborhood. From this table, we obtain difference-in-difference estimates of $\hat\delta=1.912$ for the 100m band, $0.852$ for the 200m band, and $0.754$ for the 300m band. This model is run again without neighborhood controls in Table 3. Tables 4 and 5 replicate the results in Tables 2 and 3, but instead using Log(Mean Property Value) for properties within the 100m, 200m, and 300m distance bands.

Our difference-in-difference estimates are positive and significant at the p<0.01 level in all four of the tables, for all of the specified distance bands. This analysis provides evidence of the hypothesized direction of the effect: given that our model assumptions hold, we can say with some degree of confidence that introducing affordable units into a neighborhood has a net positive effect on surrounding home values. In addition, we see that in each of the tables, the magnitude of this effect decreases (and the p-values increase) as we move from left to right across the table. This implies that as we look at properties farther and farther away from affordable developments, the magnitude of the treatment effect decreases.

Nonetheless, we have concerns that these results are providing estimates much larger than we anticipated, which may be a factor of not controlling for enough other covariates. We would expect that given the fairly small proportion of affordable properties in San Francisco - a complex, dynamic housing market - the introduction of these units would not have as large of an effect as suggested by our estimates. For example, the coefficient of $0.754$ for the 300m distance band in Table 2 implies that a developer's choice to designate a proportion of units as affordable increases median home values within 300m of that property by $100\times (e^{0.754}-1)\approx112 \%$. While it could hypothetically be possible that surrounding property values are doubling when affordable units are introduced, we believe it is more likely that unobservables or outliers are causing these coefficients to be biased upwards, or that we are mistakenly accounting for the property itself when we calculate distance band averages (which we took efforts to avoid). However, even if we did have the latter problem, we would have the same issue with the untreated group, so we are fairly confident that our coefficients are reasonable estimates of the true parameter.

@@

m1 <- lm(log.dist_100_median ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m2 <- lm(log.dist_200_median ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m3 <- lm(log.dist_300_median ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m4 <- lm(log.dist_100_median ~ treated + time + did, data = time.subset)
m5 <- lm(log.dist_200_median ~ treated + time + did, data = time.subset)
m6 <- lm(log.dist_300_median ~ treated + time + did, data = time.subset)

m7 <- lm(log.dist_100_mean ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m8 <- lm(log.dist_200_mean ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m9 <- lm(log.dist_300_mean ~ treated + time + did + as.factor(Assessor.Neighborhood.District), data = time.subset)
m10 <- lm(log.dist_100_mean ~ treated + time + did, data = time.subset)
m11 <- lm(log.dist_200_mean ~ treated + time + did, data = time.subset)
m12 <- lm(log.dist_300_mean ~ treated + time + did, data = time.subset)

stargazer(m1, m2, m3, type='latex', title="Diff-in-Diff Model for Log(Median Property Value) Outcomes, With Neighborhood Controls",
          dep.var.labels=c("Log(Median Val), 100m","Log(Median Val), 200m", "Log(Median Val), 300m"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = c("Assessor.Neighborhood.District"), omit.labels = c("Neighborhood dummies?"))

stargazer(m4, m5, m6, type='latex', title="Diff-in-Diff Model for Log(Median Property Value) Outcomes, Without Neighborhood Controls",
          dep.var.labels=c("Log(Median Val), 100m","Log(Median Val), 200m", "Log(Median Val), 300m"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = c("Assessor.Neighborhood.District"), omit.labels = c("Neighborhood dummies?"))

stargazer(m7, m8, m9, type='latex', title="Diff-in-Diff Model for Log(Mean Property Value) Outcomes, With Neighborhood Controls",
          dep.var.labels=c("Log(Mean Val), 100m","Log(Mean Val), 200m", "Log(Mean Val), 300m"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = c("Assessor.Neighborhood.District"), omit.labels = c("Neighborhood dummies?"))

stargazer(m10, m11, m12, type='latex', title="Diff-in-Diff Model for Log(Mean Property Value) Outcomes, Without Neighborhood Controls",
          dep.var.labels=c("Log(Mean Val), 100m","Log(Mean Val), 200m", "Log(Mean Val), 300m"),
          covariate.labels=c("On-Site Affordable Housing", "Post-Construction", "Diff-in-Diff Estimator", "Constant"),
          font.size="small", header=FALSE, omit = c("Assessor.Neighborhood.District"), omit.labels = c("Neighborhood dummies?"))

```{r eval=FALSE, fig.height=5, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}

@@

\newpage
We visualize the difference-in-differences results in the following plots. The top three plots reflect the trends in city-wide averages of log(median property value) at given distance bands (from upper left to right = 100m, 200m, and 300m) for the year before, year of, and year following construction of Section 415-mandated properties. The red line represents the time trend between these three periods for "treated" properties that designated affordable housing, while the blue line shows the time trend for those properties that did not designate affordable housing. The bottom three plots are identical, but using the *median* of the log(median) within-distance band property values across the city, rather than the *mean*.

![Difference-in-Differences Visualization](did.png){width=100%}

In these plots, we see a fair amount of evidence to support the claim that the parallel trends assumption holds in the data, as shown by the relatively similar slopes of the pre-construction trends for the affordable and non-affordable properties. Although this visual may not be sufficiently strong evidence to support the parallel trends assumption as it stands, our models actually run the conditional parallel trends assumption, which accounts for covariates and increases the likelihood that this assumption will hold.

As in our regression tables, we can see from these plots that the treatment tends to have a positive effect. In every case, we see that the average/median property values for the affordable properties (red line) continues to increase following construction, albeit at a slower rate than it was increasing prior to construction. On the other hand, the average/median property values for the non-affordable properties tends to decrease. This suggests that the introduction of new units into a neighborhood always tends to slows the growth of property values (all lines flatten out), but less so when affordable units are introduced. The decrease in property values following the introduction of non-affordable units (downward slope of the blue line) is consistent with theory, as well: since the supply of market-rate units is increasing without any restriction, the average value of units in the surrounding area will tend to decrease.

# Discussion

By leveraging the unique legal framework of San Francisco's affordable housing mandate, as well as the public availability of detailed micro-level spatial data, we evaluate the impact of the city's Inclusionary Housing Program on residential property values in the surrounding area. We obtain significant evidence suggesting that this policy may have a positive effect on nearby home values, contradicting its supposed aim as a tool for reducing the cost of housing. We also find evidence to support the hypothesis that the magnitude of this effect decreases with distance.

@@

A more comprehensive evaluation of the effectiveness of this program might address questions related to its effect on outcomes for individuals who are selected to participate in the program. Although individual-level data is not publicly available, the random assignment generated by the program's lottery structure could allow for a rigorous quasi-experimental approach.
